---
title: "Modelling mean streamflow from CAMELS data"
author: "Will Keenan"
format: html
---

## Lab 5

We start by loading libraries and downloading camels

```{r}
#| results: hide
#| echo: false
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
library(ggthemes)
library(ggplot2)
```

```{r}
#| results: false
#| 
cwd = getwd()
root <- 'https://zenodo.org/records/15529996/files'
download.file('https://zenodo.org/records/15529996/files/camels_attributes_v2.0.pdf?download=1',
              glue('{cwd}/data/camels_attributes_v2.0.pdf'),
              mode = 'wb')
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")
# Where the files live online ...
remote_files  <- glue('{root}/camels_{types}.txt?download=1')
# where we want to download the data ...
local_files   <- glue('{cwd}/data/camels_{types}.txt')
walk2(remote_files, local_files, download.file, quiet = TRUE)
# Read and merge data
camels <- map(local_files, read_delim, show_col_types = FALSE)
camels <- power_full_join(camels ,by = 'gauge_id')
```
## Question 1

Here we visualize how CAMELS watersheds are distributed around the country and their aridity.

```{r}
ispdf <- file.exists(glue('{cwd}/data/camels_attributes_v2.0.pdf'))
print(paste('pdf is downloaded:', ispdf))

istxts <- file.exists(local_files)
print('txts are downloaded:')
print(istxts)
```
zero_q_freq is the frequency of days with Q = 0 mm/day. It's a measure of how often the river runs dry.

```{r}
ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  annotation_borders("state", colour = "gray50") +
  geom_point(aes(color = q_mean)) +
  scale_color_gradient(low = "pink", high = "dodgerblue") +
  labs(title='Mean Q (mmd) for CAMELS gauges') +
  ggthemes::theme_map()
```

## Question 2

We model aridity, rainfall, and runoff.

```{r}
library(patchwork)
library(viridis)

left <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  annotation_borders("state", colour = "gray50") +
  geom_point(aes(color = p_mean)) +
  scale_color_viridis_c(option='magma') +
  ggthemes::theme_map() +
  labs(color='Mean Precipitation')

right <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  annotation_borders("state", colour = "gray50") +
  geom_point(aes(color = aridity)) +
  scale_color_viridis_c(option='magma', direction=-1) +
  ggthemes::theme_map() +
  labs(color='Aridity')

left / right
```
```{r}
camels |> 
  select(aridity, p_mean, q_mean) |> 
  drop_na() |> 
  cor()

# Create a scatter plot of aridity vs rainfall
ggplot(camels, aes(x = aridity, y = p_mean)) +
  # Add points colored by mean flow
  geom_point(aes(color = q_mean)) +
  # Add a linear regression line
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  # Apply the viridis color scale
  scale_color_viridis_c() +
  # Add a title, axis labels, and theme (w/ legend on the bottom)
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c() +
  # Apply log transformations to the x and y axes
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  # Apply a log transformation to the color scale
  scale_color_viridis_c(trans = "log") +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom",
        # Expand the legend width ...
        legend.key.width = unit(2.5, "cm"),
        legend.key.height = unit(.5, "cm")) + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

 ## Question 3
 Here we use Tidymodels to model mean Q with random forest, neural networks, and linear regression.
 
```{r}
 set.seed(123)
# Bad form to perform simple transformations on the outcome variable within a 
# recipe. So, we'll do it here.
camels <- camels |> 
  mutate(logQmean = log(q_mean))

# Generate the split
camels_split <- initial_split(camels, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

camels_cv <- vfold_cv(camels_train, v = 10)
```

```{r}
rec <-  recipe(logQmean ~ aridity + p_mean, data = camels_train) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())

# Prepare the data
baked_data <- prep(rec, camels_train) |> 
  bake(new_data = NULL)

# Interaction with lm
#  Base lm sets interaction terms with the * symbol
lm_base <- lm(logQmean ~ aridity * p_mean, data = baked_data)
summary(lm_base)
```

```{r}
# sanity check
summary(lm(logQmean ~ aridity + p_mean + aridity_x_p_mean, data = baked_data))

test_data <-  bake(prep(rec), new_data = camels_test)
test_data$lm_pred <- predict(lm_base, newdata = test_data)

metrics(test_data, truth = logQmean, estimate = lm_pred)

ggplot(test_data, aes(x = logQmean, y = lm_pred, colour = aridity)) +
  # Apply a gradient color scale
  scale_color_gradient2(low = "brown", mid = "orange", high = "darkgreen") +
  geom_point() +
  geom_abline(linetype = 2) +
  theme_linedraw() + 
  labs(title = "Linear Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

```{r}
# Define model
lm_model <- linear_reg() %>%
  # define the engine
  set_engine("lm") %>%
  # define the mode
  set_mode("regression")

# Instantiate a workflow ...
lm_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(lm_model) %>%
  # Fit the model to the training data
  fit(data = camels_train) 

# Extract the model coefficients from the workflow
summary(extract_fit_engine(lm_wf))$coefficients

# From the base implementation
summary(lm_base)$coefficients
```

```{r}
lm_data <- augment(lm_wf, new_data = camels_test)
dim(lm_data)
metrics(lm_data, truth = logQmean, estimate = .pred)
ggplot(lm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

```{r}
# RF model
library(baguette)
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(rf_model) %>%
  # Fit the model
  fit(data = camels_train)

rf_data <- augment(rf_wf, new_data = camels_test)
dim(rf_data)
metrics(rf_data, truth = logQmean, estimate = .pred)

ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

```{r}
wf <- workflow_set(list(rec), list(lm_model, rf_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv) 

autoplot(wf)
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

```{r}
final <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_model) |>
  fit(data = camels_train)

# Variable importance (Week 5-2 pattern)
final |>
  extract_fit_parsnip() |>
  vip::vip()
```

```{r}
# Predictions on the test set
rf_data <- augment(final, new_data = camels_test)

# Regression metrics (rmse, rsq, mae by default)
metrics(rf_data, truth = logQmean, estimate = .pred)

ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  # 1:1 line — perfect prediction
  geom_abline(linetype = 2) +
  # Actual regression line for contrast
  geom_smooth(method = "lm", col = 'red', linetype = 2, se = FALSE) +
  theme_linedraw() +
  labs(title = "Random Forest Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```


We extend the workflow_set above (which already contains the linear model and the random forest) by adding two more models.


```{r}
# XGBOOST
lr <- 1e-3
xgb_model <- boost_tree() |>
  set_engine("xgboost") |>
  set_mode("regression") |>
  translate()

# NEURAL NET
mlp_model <- mlp(hidden_units = 32, epochs = 50, activation = 'linear', learn_rate = lr, engine='nnet', mode='regression')

wf <- workflow_set(list(rec), list(lm_model, rf_model, xgb_model, mlp_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv) 

autoplot(wf)
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

The neural network has the highest accuracy and best R2. However, it is the least explainable, and only produces marginal gains over the linear regression. For this reason, I would choose linear regression. It is only slightly less accurate than random forest and the neural net and is interpretable.

## Question 4: Predict mean streamflow from camels data.

This time we consider more predictor variables

```{r}
library(skimr)
set.seed(123)

# pick variables I think might help predict mean q that don't already have mean q. I.e., we shouldn't model mean q with q5 or q95... I'm sure that would be really accurate but it's not realistic to have those variables but not q_mean.
preds <- c('q_mean', 'p_mean', 'pet_mean', 'aridity', 'frac_snow', 'elev_mean', 'slope_mean', 'frac_forest', 'lai_max')
cm <- camels |>
  select(all_of(preds), gauge_id, gauge_lat, gauge_lon)

skim(cm)
```
```{r}
library(ape)
library(ggcorrplot)
library(spatialsample)
library(sf)

dists <- as.matrix(dist(cbind(cm$gauge_lon, cm$gauge_lat)))
weights <- 1/dists
diag(weights) <- 0

# look at histograms, r, and morans I
for (col_name in preds) {
    hist(cm[[col_name]], main = paste("Histogram of", col_name))
    
    moran_result <- Moran.I(cm[[col_name]], weights, na.rm = TRUE)
    # Extract the observed I value and P-value
    i <- round(moran_result$observed, 4)
    p <- round(moran_result$p.value, 4)
    print(paste(col_name, 'i:', i, 'p:', p))
}

# Plot the corr matrix
corr <- round(cor(cm |> select(all_of(preds)), use = "complete.obs"), 2)
ggcorrplot(corr, hc.order = TRUE, type = "lower", lab = TRUE)

# add log q
cm <- cm |> 
  mutate(log_q_mean = log(q_mean))

# split
cm_split <- initial_split(cm, prop = 0.75)
cm_train <- training(cm_split)
cm_test  <- testing(cm_split)

cm_train <- st_as_sf(cm_train, coords = c('gauge_lon', 'gauge_lat'), crs=4326) |>
  sf::st_set_crs(4326)
```

```{r}
rec <-  recipe(log_q_mean ~ p_mean + frac_forest + frac_snow + pet_mean + slope_mean, data = cm_train) %>%
  # Log transform the predictor variables except p_mean (looks normal enough)
  step_log(frac_forest, frac_snow, slope_mean, pet_mean, offset = 0.01) %>%
  # Add an interaction term between aridity and p_mean
  #step_interact(terms = ~ slope_mean:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())
```

Frac_forest informs the model about transpiration. Slope informs the model about how much water produces runoff rather than is lost to deep groundwater recharge. P_mean, and pet_mean are likely the most important predictors. Aridity is just p_mean/pet_mean so we don't need that. lai_max and elevation are highly collinear with other predictors so I don't want them.

```{r}
# RF model: handles non-linearity
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# NEURAL NET: most black box but tons of predictive power
mlp_model <- mlp(hidden_units = 32, epochs = 50, activation = 'linear', learn_rate = 1E-2, engine='nnet', mode='regression')

# simplest model, captures linear relationships
lm_model <- linear_reg() %>%
  # define the engine
  set_engine("lm") %>%
  # define the mode
  set_mode("regression")
```

```{r}
simple_cv <- vfold_cv(cm_train, v = 10)
spatial_cv <- spatialsample::spatial_block_cv(cm_train)

wf <- workflow_set(list(rec), list(rf_model, mlp_model, lm_model)) %>% # , rf_model, mlp_model
  workflow_map('fit_resamples', resamples = simple_cv)

autoplot(wf)
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

```{r}
wf_spatial <- workflow_set(list(rec), list(rf_model, mlp_model, lm_model)) %>% # , rf_model, mlp_model
  workflow_map('fit_resamples', resamples = spatial_cv)
autoplot(wf_spatial)
rank_results(wf_spatial, rank_metric = 'rsq', select_best = TRUE)
```

In the non-spatial CV, the neural net model was slightly better than the neural network. The best RMSE and R2 was 0.35 and 0.91, respectively. However, this model does not consider spatial autocorrelation and so it's not as good of a representation of the model's true accuracy for data it has not seen. The spatial CV produced slightly lower accuracy, but it's a more "honest" representation of accuracy. The random forest performed slighly better for spatial CV.

```{r}
# I chose the random forest. It's accuracy is highest
rf_model2 <- rand_forest(
    mtry = tune(),
    min_n = tune(),
    trees = 500
  ) %>%
  set_engine("ranger") %>%
  set_mode("regression")

rf_grid <- grid_regular(
  mtry(range = c(1, 5)), # predictors sampled per split
  min_n(range(2, 10))
)

rf_tuning <- tune_grid(
  rf_model2,
  preprocessor = rec, 
  resamples = spatial_cv,
  grid = rf_grid,
  control = control_grid(save_pred = TRUE)
)

autoplot(rf_tuning)
show_best(rf_tuning, metric = "rsq")

best_rf_params <- select_best(rf_tuning, metric = "rsq")
print(best_rf_params)
```
The model is most sensitive to the mtry hyperparameter which is the number of predictors that will be randomly sampled at each split. The best value for this parameter is 3. The min_n parameter does not significantly change the model results. Values of 2, 6, and 10 all produced almost the same accuracy.

```{r}
final <- workflow() |>
  add_recipe(rec) |>
  add_model(finalize_model(rf_model, best_rf_params)) |>
  fit(data = cm_train)

final |>
  extract_fit_parsnip() |>
  vip::vip()
```
As expected, precipipation is the most important variable. More precipitation, more streamflow. Fraction of forest and slope are also strong predictors. High slopes promote runoff rather than infilltration, and forests transpire a lot of water. Surprisingly, PET is not very important. Fraction of snow was also not a strong predictor, although snow dominated watersheds typically have higher runoff ratios. While there is some spatial autocorrelation, none of these predictors have Moran's I values above 0.5. This combined with the use of spatial CV prevents the model from using geographic proxies rather than true relationships.

```{r}
final <- workflow() |>
  add_recipe(rec) |>
  add_model(finalize_model(rf_model, best_rf_params))

final <- finalize_workflow(final, best_rf_params)
final <- last_fit(final, split=cm_split)
metrics <- final %>% collect_metrics()

rmse <- round(exp(metrics$.estimate[1]),2)
rsq <- round(metrics$.estimate[2],2)

cm_test <- final %>% 
  extract_workflow() %>% 
  augment(new_data = testing(cm_split))

# untransform it. Regular ppl don't want to see log flow.

cm_test$q_mean_pred = exp(cm_test$.pred)
cm_test |>
  ggplot(aes(x = q_mean_pred, y = q_mean, color=aridity)) + 
  scale_color_viridis_c(option = "magma") +
  geom_point(alpha = 0.5) +
  geom_abline(lty = 2, color = "red") + 
  labs(
    title = "Observed vs. Predicted Q",
    x = "Predicted Q (mmd)",
    y = "Observed Q (mmd)",
    subtitle = paste('RMSE:', rmse, 'mmd', 'R2:', rsq)
  ) +
  coord_obs_pred() 
```
The model performs very well. An R2 of 0.91 is exceptionally high. The test set performed slightly better than the spatial CV. This is surprising, but it could be due to our hyperparameter tuning or just random chance. The model has very low bias as the residuals are scattered evenly around the 1:1 line. There is one notable outlier where Q is underpredicted.



