---
title: "lab4"
author: "Will Keenan"
format: html
---

```{r}
library(rstac) # STAC API
library(terra) # Raster Data handling
library(sf) # Vector data processing
library(mapview) # Rapid Interactive visualization
library(AOI)
library(tidyr)
library(dplyr)


linn <- aoi_get(state = "IA", county = "Linn")
linn_bbox <- linn |>
  st_transform(4326) |>
  st_bbox()

(stac_query <- stac("https://planetarycomputer.microsoft.com/api/stac/v1"))
bands <- c('coastal', 'blue', 'green', 'red', 'nir08', 'swir16')

scenes <- stac_query |>
  stac_search(
    collections = "landsat-c2-l2",
    bbox = as.numeric(linn_bbox),
    datetime = "2016-09-26/2016-09-29"
  ) |>
  get_request() |>
  items_fetch()

print(scenes)

signed_scenes <- rstac::items_sign(
  scenes,
  rstac::sign_planetary_computer()
)

```


```{r}
cwd <- getwd()
out <- file.path(cwd, 'data/lab4')

if (length(list.files(out)) > 0) {
  message(out, "' already contains files. Skipping download.")
} else {
  message("Directory is empty. Starting download...")
  assets_download(signed_scenes, output_dir = out, overwrite = FALSE)
}
```


```{r}
ims <- file.path(out, r"(landsat-c2/level-2/standard/oli-tirs/2016/025/031/LC08_L2SP_025031_20160926_20200906_02_T1)")
                 
files <- list.files(path = ims, pattern = "\\.TIF$", full.names = TRUE)
print(files)

im <- rast(files)
bandnames <- names(im)

# rename bands
clean_names <- sub(".*_", "", bandnames)
print(clean_names)
names(im) <- clean_names
```
View images

```{r}
# true color
plotRGB(im, r = "B4", g = "B3", b = "B2", stretch = "lin")
# false color NIR, R, G
plotRGB(im, r = "B5", g = "B4", b = "B3", stretch = "lin")
# false color SWIR, NIR, R
plotRGB(im, r = "B6", g = "B5", b = "B4", stretch = "lin")
# false color 
plotRGB(im, r = "B7", g = "B6", b = "B4", stretch = "lin")
```

True Color images are best for seeing the landscape in a familiar way. It helps us identify objects because we are used to seeing them in RGB. False color 432 images highlight vegetation because NIR is highly reflective in plants.
False color 654 highlights water features by making them very dark relative to everything else. False color 764 highlights urban infrastructure.

```{r}
# NDVI
index <- (im$B5 - im$B4) / (im$B5 + im$B4)
names(index) <- "NDVI"
im <- c(im, index)
# NDWI
index <- (im$B3 - im$B5) / (im$B3 + im$B5)
names(index) <- "NDWI"
im <- c(im, index)
# MNDWI
index <- (im$B3 - im$B6) / (im$B3 + im$B6)
names(index) <- "MNDWI"
im <- c(im, index)
# water ratio index
index <- (im$B4 + im$B3) / (im$B5 + im$B6)
names(index) <- "WRI"
im <- c(im, index)
# simple water index
index <- 1 / ((im$B2 - im$B6)**0.5)
names(index) <- "SWI"
im <- c(im, index)

```
```{r}
# plot them
color <- colorRampPalette(c("blue", "white", "red"))
plot(im$SWI, col = color(256), main = "SWI")
plot(im$NDVI, col = color(256), main = "NDVI")
plot(im$MNDWI, col = color(256), main = "MNDWI")
plot(im$NDWI, col = color(256), main = "NDWI")
plot(im$WRI, col = color(256), main = "WRI")
```

Unlike the other 4 indices, NDVI is designed to highlight vegetation, not water. However, water has a very low NDVI so it's still outlines the river quite well. Out of the 4 water indices, MNDWI does the best job of clearly delineating the river.

```{r}
NDVIt <- ifel(im$NDVI < 0, 1, 0)
NDWIt <- ifel(im$NDWI > 0, 1, 0)
MNDWIt <- ifel(im$MNDWI > 0, 1, 0)
WRIt <- ifel(im$WRI > 1, 1, 0)
SWIt <- ifel(im$SWI < 5, 1, 0)
```

```{r}
stack <- c(NDVIt, NDWIt, MNDWIt, WRIt, SWIt)
set.names(stack, c('NDVI', 'NDWI', 'MNDWI', 'WRI', 'SWI'))

plot(stack$NDVI, main='NDVI')
plot(stack$NDWI, main='NDWI')
plot(stack$MNDWI, main='MNDWI')
plot(stack$WRI, main='WRI')
plot(stack$SWI, main='SWI')
```

MNDWI has the most sensitivity as there are a lot of pixels outside of the main river network. SWI has a lot of NAs. Overall, they all look very similar.


```{r}
# remove SWI because there are so many NAs
#stack <- subset(stack, "SWI", negate = TRUE)

vals <- values(stack)
clean_idx <- which(complete.cases(vals))
k_results <- kmeans(vals[clean_idx, ], centers = 5, iter.max = 100, nstart = 5)

clust_raster <- stack$NDVI
values(clust_raster) <- NA
clust_raster[clean_idx] <- k_results$cluster

levels(clust_raster) <- data.frame(ID = 1:5, cluster = paste0("C", 1:5))
```

```{r}
plot(clust_raster, main = "K-Means Cluster Map")
```

```{r}
ndwi_values <- values(stack$NDWI)
cluster_values <- values(clust_raster)

flood_tab <- table(ndwi_values, cluster_values)

print(flood_tab)

```

```{r}
print(paste('flood class:', which.max(flood_tab[2,])))
class <- which.max(flood_tab[2,])

names(clust_raster) <- "kmeans"
clust_raster <- ifel(clust_raster==class, 1, 0)
stack <- c(stack, clust_raster)
# see which has the most flooded pixels
NDVI_sum <- global(stack$NDVI, fun = "sum", na.rm = TRUE)
NDWI_sum <- global(stack$NDWI, fun = "sum", na.rm = TRUE)
MNDWI_sum <- global(stack$MNDWI, fun = "sum", na.rm = TRUE)
WRI_sum <- global(stack$WRI, fun = "sum", na.rm = TRUE)
SWI_sum <- global(stack$SWI, fun = "sum", na.rm = TRUE)

print(paste('Total flooded pixels per index: NDVI:', NDVI_sum, 'NDWI:', NDWI_sum, 'MNDWI:', MNDWI_sum, 'WRI:', WRI_sum, 'SWI:', SWI_sum))
```

```{r}
# sum them all up
sum_floods <- app(stack, fun = sum, na.rm=TRUE)
plot(sum_floods, col = blues9, main='Sum of each flood mapping raster')
```

The cells are all integers between 0 and 6, as expected.