---
title: "lab4"
author: "Will Keenan"
format: html
---

```{r}
library(rstac) # STAC API
library(terra) # Raster Data handling
library(sf) # Vector data processing
library(mapview) # Rapid Interactive visualization
library(AOI)
library(tidyr)
library(dplyr)


linn <- aoi_get(state = "IA", county = "Linn")
linn_bbox <- linn |>
  st_transform(4326) |>
  st_bbox()

(stac_query <- stac("https://planetarycomputer.microsoft.com/api/stac/v1"))
bands <- c('coastal', 'blue', 'green', 'red', 'nir08', 'swir16')

scenes <- stac_query |>
  stac_search(
    collections = "landsat-c2-l2",
    bbox = as.numeric(linn_bbox),
    datetime = "2016-09-26/2016-09-29"
  ) |>
  get_request() |>
  items_fetch()

print(scenes)

signed_scenes <- rstac::items_sign(
  scenes,
  rstac::sign_planetary_computer()
)

```


```{r}
#cwd <- getwd()
out <- file.path(cwd, 'data/lab4')
assets_download(signed_scenes, output_dir = out, overwrite = TRUE)
```


```{r}
ims <- file.path(out, r"(landsat-c2/level-2/standard/oli-tirs/2016/025/031/LC08_L2SP_025031_20160926_20200906_02_T1)")
                 
files <- list.files(path = ims, pattern = "\\.TIF$", full.names = TRUE)
print(files)

im <- rast(files)
bandnames <- names(im)

# rename bands
clean_names <- sub(".*_", "", bandnames)
print(clean_names)
names(im) <- clean_names
```
View images

```{r}
# true color
plotRGB(im, r = "B4", g = "B3", b = "B2", stretch = "lin")
# false color NIR, R, G
plotRGB(im, r = "B5", g = "B4", b = "B3", stretch = "lin")
# false color SWIR, NIR, R
plotRGB(im, r = "B6", g = "B5", b = "B4", stretch = "lin")
# false color 
plotRGB(im, r = "B7", g = "B6", b = "B4", stretch = "lin")
```

True Color images are best for seeing the landscape in a familiar way. It helps us identify objects because we are used to seeing them in RGB. False color 432 images highlight vegetation because NIR is highly reflective in plants.
False color 654 highlights water features by making them very dark relative to everything else. False color 764 highlights urban infrastructure.

```{r}
# NDVI
index <- (im$B5 - im$B4) / (im$B5 + im$B4)
names(index) <- "NDVI"
im <- c(im, index)
# NDWI
index <- (im$B3 - im$B5) / (im$B3 + im$B5)
names(index) <- "NDWI"
im <- c(im, index)
# MNDWI
index <- (im$B3 - im$B6) / (im$B3 + im$B6)
names(index) <- "MNDWI"
im <- c(im, index)
# water ratio index
index <- (im$B4 + im$B3) / (im$B5 + im$B6)
names(index) <- "WRI"
im <- c(im, index)
# simple water index
index <- 1 / (im$B2 - im$B6)**0.5
names(index) <- "SWI"
im <- c(im, index)


```
```{r}
# plot them
color <- colorRampPalette(c("blue", "white", "red"))
plot(im$SWI, col = color(256), main = "SWI")
plot(im$NDVI, col = color(256), main = "NDVI")
plot(im$MNDWI, col = color(256), main = "MNDWI")
plot(im$NDWI, col = color(256), main = "NDWI")
plot(im$WRI, col = color(256), main = "WRI")
```

Unlike the other 4 indices, NDVI is designed to highlight vegetation, not water. However, water has a very low NDVI so it's still outlines the river quite well. Out of the 4 water indices, MNDWI does the best job of clearly delineating the river.

```{r}
NDVIt <- ifel(im$NDVI < 0, 1, 0)
NDWIt <- ifel(im$NDWI > 0, 1, 0)
MNDWIt <- ifel(im$MNDWI > 0, 1, 0)
WRIt <- ifel(im$WRI > 1, 1, 0)
SWIt <- ifel(im$SWI < 5, 1, 0)
```

```{r}
stack <- c(NDVIt, NDWIt, MNDWIt, WRIt, SWIt)
set.names(stack, c('NDVI', 'NDWI', 'MNDWI', 'WRI', 'SWI'))

plot(stack$NDVI, main='NDVI')
plot(stack$NDWI, main='NDWI')
plot(stack$MNDWI, main='MNDWI')
plot(stack$WRI, main='WRI')
plot(stack$SWI, main='SWI')
```