---
title: "Lab 3: MAUP in Practice — Tessellations and Point-in-Polygon"
subtitle: "National Dam Inventory (NID): From Counts to Local Clustering"
format:
  html:
    code-fold: true
    toc: true
editor: 
  markdown: 
    wrap: 72
---

# Background

In this lab we will explore the impacts of tessellated surfaces and the
modifiable areal unit problem (MAUP) using the National Dam Inventory
maintained by the United States Army Corps of Engineers. The work
requires writing reusable functions and careful consideration of feature
aggregation/simplification, spatial joins, and data visualization. The
end goal is to visualize the distribution of dams and their purposes
across the country.

### Libraries

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

# Data Science
library(tidyverse) # data wrangling
library(sf)        # Working with vector data
library(units)     # manage your units
# Data
#remotes::install_github("mikejohnson51/AOI") # get US counties
library(AOI) # county boundaries

# Visualization
library(gghighlight) # ggplot conditional highlighting
library(knitr) # table generation
library(kableExtra) # making tables pretty
```

# Question 1:

Here we will prepare five tessellated surfaces from CONUS and write a
function to plot them in a descriptive way.

### Step 1.1

First, we need a spatial file of CONUS counties. For future area
calculations we want these in an equal area projection (`EPSG:5070`).

```{r}
counties <- AOI::aoi_get(state = "conus", county = "all")
counties <- st_transform(counties, crs=5070)
```

### Step 1.2

For triangle based tessellations we need point locations to serve as our
"anchors".

To achieve this:

-   generate county centroids using `st_centroid`

-   Since, we can only tessellate over a feature we need to *combine* or
    *union* the resulting 3,108 `POINT` features into a single
    `MULTIPOINT` feature

-   Since these are point objects, the difference between union/combine
    is mute

```{r}
ctrds <- st_centroid(counties)
ctrds <- st_union(ctrds)
```

### Step 1.3

Tessellations/Coverage's describe the **extent** of a region with
geometric shapes, called **tiles**, with *no* overlaps or gaps.

**Tiles** can range in *size*, *shape*, *area* and have different
methods for being created.

Some methods generate triangular tiles across a set of defined points
(e.g. `Voronoi` and Delaunay triangulation)

Others generate equal area tiles over a known extent (`st_make_grid`)

For this lab, we will create surfaces of CONUS using using 4 methods, 2
based on an extent and 2 based on point anchors:

**Tessellations** :

-   `st_voronoi`: creates a Voronoi tessellation

-   `st_triangulate`: triangulates a set of points (not constrained)

**Coverages**:

-   `st_make_grid`: creates a square grid covering the geometry of an sf
    or sfc object

-   `st_make_grid(square = FALSE)`: creates a hexagonal grid covering
    the geometry of an sf or sfc object

-   The side of coverage tiles can be defined by a cell resolution or a
    specified number of cell in the X and Y direction

```{r}
fix_tes <- function(sfc){
  sfc_fix <- sfc |>
    st_collection_extract("POLYGON") |>
    st_sf() |>
    mutate(id = seq(len = n())) |>
    st_cast()
    return(sfc_fix)
}

vrni <- st_voronoi(ctrds)
vrni <- fix_tes(vrni)

tri <- st_triangulate(ctrds)
tri <- fix_tes(tri)

grid <- st_make_grid(ctrds, n=70)
grid <- fix_tes(grid)

hgrid <- st_make_grid(ctrds, square=FALSE, n=70)
hgrid <- fix_tes(hgrid)
```

### Step 1.4

If you plot the above tessellations you'll see that the **Voronoi**,
**Delaunay triangulation**, **hexagon**, and **square grids** all
produce regions far beyond the boundaries of CONUS.

We need to cut these boundaries to CONUS borders.

```{r, eval = TRUE}
conus <- counties |>
  st_union() |>
  st_sf()
```

### Step 1.5

With a single feature boundary, we must carefully consider the
complexity of the geometry. Remember, the more points our geometry
contains, the more computations needed for spatial predicates our
differencing. For a task like ours, we do not need a finely resolved
coastal boarder.

```{r, eval = TRUE}
library(rmapshaper)
boundary_simplified <- ms_simplify(conus, keep=0.05)
```

```{r}
#| eval: false
# Check: Is the simplified boundary valid?
st_is_valid(boundary_simplified)

# Check: Does it still contain all of CONUS?
st_bbox(boundary_simplified)  # Should cover roughly [-125, 24] to [-66, 49]

# Check: Visual comparison
plot(conus, main = "Original")
plot(boundary_simplified, main = "Simplified")
```

```{r}
library(mapview)
b4 <- npts(conus)
after <- npts(boundary_simplified)

# rename so we use it in the future
conus <- boundary_simplified

print(paste('Points before simplification:', b4, 'Points removed:', b4-after))
```

This operation greatly reduces the amount of computatation required for
any future steps on the boundary.

-   Finally, crop **all four grids** to the CONUS boundary—but using
    different methods:
    -   **Voronoi & Delaunay** (geometry-based tessellations): Use
        `st_intersection()` to *cut* these geometries at the CONUS
        boundary. These tessellations create boundaries that extend well
        beyond CONUS; intersection actually trims them.
    -   **Hexagon & Square grids** (regular grids): Use `st_filter()` to
        keep only tiles that touch CONUS. These grids are already
        regularly aligned; filtering just discards edge tiles without
        modifying them.

```{r}
vrni <- st_intersection(vrni, conus)
tri <- st_intersection(tri, conus)

grid <- st_filter(grid, conus)
hgrid <- st_filter(hgrid, conus)
```

::: callout-tip
## Why Different Cropping Methods?

**The distinction matters for your spatial analysis:**

-   **`st_intersection()` on Voronoi/Delaunay**: Produces
    edges/boundaries that align precisely with the CONUS border. Tiles
    at the edge are cut, not discarded. This is geometrically accurate
    but creates irregular edge shapes.

-   **`st_filter()` on hexagon/square grids**: Keeps only complete
    regular tiles that touch CONUS, discarding edge tiles. The boundary
    is a stair-step pattern following the regular grid.

**In Q3**, when you compare these tessellations, you'll notice that
Voronoi and Delaunay have denser tile coverage (e.g., \~2800–6200 tiles)
because they extend to the exact boundary. Hexagons and squares have
fewer tiles (e.g., \~1900–2000) because they're bound to a regular grid.
**This difference affects spatial statistics**: denser tessellations
reveal finer-grained patterns; coarser tessellations show broader
regional structure. This is part of the Modifiable Areal Unit Problem
(MAUP).
:::

### Step 1.6

```{r}
amap <- function(sf_obj1, sf_obj2, title = "map") {
  caption <- nrow(sf_obj2)
  ggplot() +
    # CONUS outline)
    geom_sf(data = sf_obj1, fill = "white", color = "navy", linewidth = 0.5) +

    # tesselation
    geom_sf(data = sf_obj2, fill = "transparent", color = "blue", alpha = 0.4) +
    theme_void() +
    labs(
      title = title,
      caption = paste('# of features:', caption)
    )
}
```

### Step 1.7

Use your new function to plot each of your tessellated surfaces and the
original county data (5 plots in total):

```{r}
amap(vrni, boundary_simplified, 'Voroni polygons')
amap(tri, boundary_simplified, 'Triangles')
amap(grid, boundary_simplified, 'Square grid')
amap(hgrid, boundary_simplified, 'Hexagonal grid')
amap(counties, boundary_simplified, 'Original Counties')
```

# Question 2:

In this question, we will write out a function to summarize our
tessellated surfaces.

### Step 2.1

```{r}
sum_tess <- function(arg1, arg2) {
  areas <- st_area(arg1) |> 
    set_units("km^2") |> 
    drop_units()
  
  df_summary <- data.frame(
    description = arg2,
    feature_count = round(length(st_geometry(arg1))),
    mean_area_km2 = round(mean(areas)),
    sd_area = round(sd(areas)),
    total_area_km2 = round(sum(areas))
  )
  
  return(df_summary)
}
```

### Step 2.2

### Step 2.3

```{r}
smry <- bind_rows(
  sum_tess(tri, "triangulation"),
  sum_tess(vrni, "voronoi"),
  sum_tess(grid, 'grid'),
  sum_tess(hgrid, 'hgrid')
)
```

### Step 2.4

```{r, echo = FALSE}
library(flextable)
flextable(smry)
```

### Step 2.5

Comment on the traits of each tessellation. Be specific about how these
traits might impact the results of a point-in-polygon analysis in the
contexts of the modifiable areal unit problem and with respect
computational requirements.

The grid and hexagonal grid are uniform tessellations; all the polygons
are the exact same shape and size. This equally weights all areas of the
US. The voroni polygons and triangluations vary in size based on the
density of counties. These tesselations are generally correlated with
population density because they are derived from county centroids which
are more dense where there are more people.

------------------------------------------------------------------------

# Question 3: Dam Distribution Analysis & Tessellation Selection

For the remainder of this lab we will analyze the distributions of these
dams using point-in-polygon analysis.

### Step 3.1

```{r}
#| message: false
library(janitor)
# Prefer a local NID CSV named `nation.csv`; otherwise download to that filename
nid_url <- "https://nid.sec.usace.army.mil/api/nation/csv"
dir.create("data", showWarnings = FALSE)
nid_dest <- "data/nation.csv"
if (file.exists(nid_dest)) {
  message("Using local NID file: ", nid_dest)
} else {
  tryCatch({
    message("Downloading NID CSV to: ", nid_dest)
    download.file(nid_url, nid_dest, quiet = FALSE, mode = "wb")
  }, error = function(e) {
    message("NID download failed: ", e$message)
  })
}

# read the CSV (will error if download failed and file missing)
if (!file.exists(nid_dest)) stop("NID CSV not found at ", nid_dest)
dams <- readr::read_csv(nid_dest, skip =1, show_col_types = FALSE) |> 
  janitor::clean_names()

usa <- AOI::aoi_get(state = "conus") %>% 
  st_union() %>% 
  st_transform(5070)

dams2 <- dams %>% 
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(5070) %>% 
  st_filter(usa)
```

### Step 3.2

```{r}
pinp <- function(pts, polys, id_col, colname = 'point_count') {
  
  if (st_crs(polys) != st_crs(pts)) {
    message("Transforming crs's to match")
    pts <- st_transform(pts, st_crs(polys))
  }

  joined <- st_join(pts, polys, join = st_intersects)

  counts <- joined |>
    st_drop_geometry() |>
    count(.data[[id_col]], name = colname)

  result <- polys |>
    left_join(counts, by = id_col) |>
    mutate(point_count = coalesce(point_count, 0)) # Replace NAs with 0
  return(result)
}
```

### Step 3.3

```{r}
counties$id <- counties$fip_code
jgrid <- pinp(dams2, grid, 'id')
jhgrid <- pinp(dams2, hgrid, 'id')
jvrni <- pinp(dams2, vrni, 'id')
jtri <- pinp(dams2, tri, 'id')
jcntys <- pinp(dams2, counties, 'id')

```

### Step 3.4

```{r}
amap2 <- function(sf_obj1, sf_obj2, title = "map", fillvar = 'point_count') {
  caption <- sum(sf_obj2$point_count)
  plt <- ggplot() +
    # CONUS outline)
    geom_sf(data = sf_obj1, fill = "white", color = "navy", linewidth = 0.5) +

    # tesselation
    geom_sf(data = sf_obj2, aes(fill = get(fillvar)), alpha = 0.4) +
    scale_fill_viridis_c(option = "plasma") +
    theme_void() +
    labs(
      title = title,
      caption = paste(caption, 'total dams'),
      fill = fillvar
    )
  return(plt)
}
```

### Step 3.5

### Step 3.6

**Policy Analysis: Tessellation and Decision-Making**

Now that you have dam counts across all five tessellation methods,
compare the results. Specifically:

**a.** Create a side-by-side visualization (faceted or small multiples)
showing dam clustering under each of the five tessellations. Use
`patchwork` or `cowplot` to arrange the 5 plots you created in Step 3.5
for direct visual comparison.

Your answer should be **3–4 paragraphs** and grounded in the actual
visualizations and patterns you observe from your tessellation
comparisons.

```{r}
library(patchwork)
tr <- amap2(conus, jgrid, 'square grid')
tl <- amap2(conus, jhgrid, 'hexagonal grid')
br <- amap2(conus, jvrni, 'voronoi polygons')
bl <- amap2(conus, jtri, 'triangles')
bm <- amap2(conus, jcntys, 'counties')

combo <- (tl + tr) / (bl + bm + br)

combo + 
  plot_annotation(
    title = "Comparison of Tesselations",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))
  )
```

```{r}
# I cant see any detail if they are all in the same plot
amap2(conus, jgrid, 'square grid')
amap2(conus, jhgrid, 'hexagonal grid')
amap2(conus, jvrni, 'voronoi polygons')
amap2(conus, jtri, 'triangles')
amap2(conus, jcntys, 'counties')
```

**b.** Identify **2–3 geographic regions** where tessellation choice
produces substantially *different* patterns (e.g., one method shows a
clear hotspot while another shows diffuse distribution).

The voronoi polygons, triangles, and counties create a hotspot in
Northwestern Montana that is not present in the grid tesselations. This
is because the counties in Montana are large, so many dams are contained
in one or two polygons. This creates a hotspot because the density of
dams per polygon is high relative to the rest of the polygons.

In contrast, Appalachia is a hotspot in the gridded tessellations, but
not in the irregular tessellations. This is because counties are very
dense and small in this part of the country, and so most polygons in the
irregular tessellations have very few dams. In the uniform
tessellations, this area of the country has the most dams per area, and
so a hotspot is visible.

**c.** **Which single tessellation would you recommend to a dam
operations manager seeking to understand geographic clustering?**

The square gridded tessellation provides the best, most interpretable
visualization for geographic clustering. Unlike Voronoi, triangles, or
raw county boundaries, this tessellation is unbiased by the history of
county boundary creations, conflicts, politics, population density, and
natural features such as rivers and mountain ranges. It gives us a clear
view of the country, allowing us to answer "where are dams most
clustered?" without needing to consider how the county boundaries are
affecting the visible patterns.

The irregular tessellations are not valid to understand geographic
clustering. I cannot think of a use case for the Voronoi or triangle
tessellations in the dam density context. The county polygon
tessellation may be useful if you are specifically interested in which
counties have many dams and if counties had jurisdiction over dams, but
they typically don't.

The hexagonal tessellation provides no clear advantage over the square
and has the disadvantages of being unable to be converted into a raster
format and being less common. The square tessellation best supports
management decisions across many jurisdictional boundaries. However, we
could adjust the size of our square grid based on the specific use case,
as this may impact where clusters are visible.

# Question 4

The NID provides a comprehensive data dictionary
[here](https://files.hawaii.gov/dbedt/op/gis/data/nid_dams_data_dictionary.htm#Purposes).
In it we find that dam purposes are designated by a character
[code](https://files.hawaii.gov/dbedt/op/gis/data/nid_dams_data_dictionary.htm#Purposes).

```{r}
#| echo: false
#| eval: true

nid_classifier = data.frame(matrix(c('I' , 'Irrigation',
'H' , 'Hydroelectric',
'C' , 'Flood Control',
'N' , 'Navigation',
'S' , 'Water Supply',
'R' , 'Recreation',
'P' , 'Fire Protection',
'F' , 'Fish and Wildlife',
'D' , 'Debris Control',
'T' , 'Tailings',
'G' , 'Grade Stabilization',
'O' , 'Other'), ncol = 2, byrow = T)) %>% 
  setNames(c("abbr", "purpose"))
```

```{r}
#| echo: false
#| eval: true

knitr::kable(nid_classifier,              
        caption = "NID 2019: Dam Purposes",
        format.args = list(big.mark = ",")) %>% 
    kableExtra::kable_styling("striped", full_width = TRUE, font_size = 14)
```

-   In the data dictionary, we see a dam can have *multiple* purposes.

-   In these cases, the purpose codes are concatenated in order of
    decreasing importance. For example, `SCR` would indicate the primary
    purposes are *Water Supply*, then *Flood Control*, then
    *Recreation.*

-   A standard summary indicates there are over 400 unique combinations
    of dam purposes:

```{r, echo = TRUE, eval = TRUE}
unique(dams2$purposes) %>% length()
```

```{r}
#| echo: false
#| eval: false
# create a vector of all characters in your purpose and unlist 
dam_freq <- strsplit(dams2$purposes, split = ";") %>%
  unlist() %>% 
  table() %>% 
  as.data.frame() %>% 
  setNames(c("abbr", "count")) %>% 
  left_join(nid_classifier) %>% 
  mutate(lab = paste0(purpose, "\n(", abbr, ")"))
```

The result of this would indicate:

```{r}
#| echo: false
#| eval: false

ggplot() + 
  geom_col(data = dam_freq, aes(x = reorder(lab, -count), y = count, fill = abbr), alpha = .8) + 
  labs(x = "", y = "# Dams", title = "Number of Dams serving each purpose")  +
  scale_fill_viridis_d()+ 
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, vjust = .9, hjust = 1)) + 
  theme(legend.position = 'none')
```

### Step 4.1

Your task is to create point-in-polygon counts for at *least* 4 of the
above dam purposes:

```{r}
ps <- c('Recreation', 'Flood Risk Reduction', 'Fire Protection, Stock, Or Small Fish Pond', 'Irrigation')
re <- paste(ps, collapse = "|")

dams3 <- dams2 |>
  filter(!is.na(purposes)) |>
  filter(str_detect(re, purposes))

jgrid2 <- pinp(dams3, grid, 'id')
```

### Step 4.2

-   Now use your plotting function from Q3 to map these counts on the
    **cropped tessellation**.

```{r}
meanplussd <- mean(jgrid2$point_count) + sd(jgrid2$point_count)

amap2(conus, jgrid2, 'n') +
  gghighlight(point_count>meanplussd)
```

### Step 4.3

Comment on the geographic distribution of dams you found. Does it make
sense? How might the tessellation you chose impact your findings? How
does the distribution of dams coincide with other geographic factors
such as river systems, climate, ect?

The hotspots in the Great Plains and Black Hills regions are a little
surprising given the aridity and low population density. However, this
dataset is a raw count of dams, so these hotspots may just be a lot of
really small dams mostly for recreation purposes. I also doubt that
these dams were truly built for recreation, although that may be their
main purpose in the modern era. The uncontrolled dam building of the
50s, 60s, and 70s led to a lot of useless dams being built. The other
hotspots are generally clustered where there is a lot of rainfall,
rivers, and people.

### Step 5.1

**Aggregate Dam Attributes by Tile**

Using your **cropped tessellation from Q1.4** (your recommended grid
from Q3, clipped to CONUS boundary), compute **mean and variance** of
key dam attributes within each tile.

```{r}
# Suggested workflow (write your own implementation):
# 1) spatial join dams to your cropped tessellation
# 2) group by tile id and summarize n_dams, mean_year_completed, mean_surface_area
# 3) join summaries back to geometry and preserve zero-dam tiles
# 4) inspect distributions before mapping
joined <- st_join(grid, dams2) |>
  group_by(id) |>
  summarize(n_dams = n_distinct(dam_name, na.rm =TRUE), mean_y_c = mean(year_completed, na.rm=TRUE), mean_sa = mean(surface_area_acres, na.rm=TRUE)) |>
  select(id, n_dams, mean_y_c, mean_sa)

# do it for counties to support Q6
jc <- st_join(counties, dams2) |>
  group_by(id) |>
  summarize(n_dams = n_distinct(dam_name, na.rm =TRUE), mean_y_c = mean(year_completed, na.rm=TRUE), mean_sa = mean(surface_area_acres, na.rm=TRUE)) |>
  select(id, n_dams, mean_y_c, mean_sa)

```

**Produce visualizations** showing spatial patterns:

-   Map 1: Mean dam age by tile (use viridis scale)
-   Map 2: Mean surface area by tile
-   Map 3: Dam count by tile (you already have this from Q3)

Do you see geographic clustering of old vs. new dams? Large vs. small
dams?

```{r}
joined$age <- 2026 - joined$mean_y_c
jc$age <- 2026 - jc$mean_y_c

amap2(conus, joined, title='Mean Dam Age', fillvar='age')
amap2(conus, joined, title='Mean Surface Area', fillvar='mean_sa')
amap2(conus, joined, title='Number of dams', fillvar='n_dams')
```

Dams are slighly older out West where irrigation farming is the only way
to grow crops.

### Step 5.2

```{r}
#install.packages('sfdep')
library(sfdep)
joined$geometry <- joined$st_collection_extract.sfc...POLYGON..
# Suggested workflow (write your own implementation):
# 1) build vectors for count, age, and size attributes by tile
# 2) handle NAs explicitly before matrix multiplication
# 3) compute spatial lag as W_std %*% attribute
# 4) correlate each attribute with its lag
# 5) report values and classify as weak/moderate/strong with your own thresholds
joined <- joined |>
  mutate(mean_y_c2 = ifelse(is.na(mean_y_c), mean(mean_y_c, na.rm=TRUE), mean_y_c),
         mean_sa2 = ifelse(is.na(mean_sa), mean(mean_sa, na.rm=TRUE), mean_sa),
         n_dams2 = ifelse(is.na(n_dams), mean(n_dams, na.rm=TRUE), n_dams),
         age2 = ifelse(is.na(age), mean(age, na.rm=TRUE), age),
         neighbors = st_contiguity(geometry),
         weights = st_weights(neighbors),
         w_std_n_dams = st_lag(n_dams2, neighbors, weights),
         w_std_age = st_lag(age2, neighbors, weights),
         w_std_sa = st_lag(mean_sa2, neighbors, weights))
        
agecor <- cor(joined$age, joined$w_std_age, use="complete.obs")
damscor <- cor(joined$n_dams, joined$w_std_n_dams, use="complete.obs")
sacor <- cor(joined$mean_sa, joined$w_std_sa, use="complete.obs")

print(paste('age cor:', agecor, 'dams cor:', damscor, 'sa cor:', sacor))

# redo for counties
jc <- jc |>
  mutate(mean_y_c2 = ifelse(is.na(mean_y_c), mean(mean_y_c, na.rm=TRUE), mean_y_c),
         mean_sa2 = ifelse(is.na(mean_sa), mean(mean_sa, na.rm=TRUE), mean_sa),
         n_dams2 = ifelse(is.na(n_dams), mean(n_dams, na.rm=TRUE), n_dams),
         age2 = ifelse(is.na(age), mean(age, na.rm=TRUE), age),
         neighbors = st_contiguity(geometry),
         weights = st_weights(neighbors),
         w_std_n_dams = st_lag(n_dams2, neighbors, weights),
         w_std_age = st_lag(age2, neighbors, weights),
         w_std_sa = st_lag(mean_sa2, neighbors, weights))

```

**Summarize in a table**:

```{r}
table = data.frame(matrix(c('age' , round(agecor,2), 'moderate',
'n_dams', round(damscor,2), 'high', 'surface_area', round(sacor,2), 'none'),
ncol = 3, byrow = T)) %>% 
  setNames(c("name", "correlation", "qualitative evaluation"))

flextable(table)
```

### Step 5.3

**Global Question:** Do neighboring tiles resemble each other overall,
or is the pattern weak?

Create a single **Moran's I scatterplot** for one dam attribute from
Step 5.2 (for example, mean year completed or dam count). Use
standardized values on both axes: the attribute on the x-axis and the
spatial lag on the y-axis. The plot should show whether points
concentrate along the diagonal or remain widely dispersed.

```{r}
# Suggested workflow (write your own implementation):
# 1) standardize the selected attribute and its spatial lag
# 2) assign HH / LL / HL / LH quadrants by sign of (z, lag_z)
# 3) produce a Moran scatterplot with dashed axes at 0
# 4) report quadrant counts and interpret what dominates

add_quadrants <- function(data, var1, var2, prefix = "age") {
  data |>
    mutate(
      "{prefix}_z" := as.numeric(scale({{var1}})),
      "{prefix}_z_std" := as.numeric(scale({{var2}})),
      "{prefix}_quad" := case_when(
        .data[[paste0(prefix, "_z")]]  < 0 & .data[[paste0(prefix, "_z_std")]]  < 0 ~ "LL",
        .data[[paste0(prefix, "_z")]]  < 0 & .data[[paste0(prefix, "_z_std")]] >= 0 ~ "LH",
        .data[[paste0(prefix, "_z")]] >= 0 & .data[[paste0(prefix, "_z_std")]] >= 0 ~ "HH",
        TRUE                                                                       ~ "HL"
      )
    )
}

joined <- joined |>
  add_quadrants(age, w_std_age, prefix = "age") |>
  add_quadrants(n_dams, w_std_n_dams, prefix = "n_dams") |>
  add_quadrants(mean_sa, w_std_sa, prefix = "sa")

jc <- jc |>
  add_quadrants(age, w_std_age, prefix = "age") |>
  add_quadrants(n_dams, w_std_n_dams, prefix = "n_dams") |>
  add_quadrants(mean_sa, w_std_sa, prefix = "sa")
  

ggplot(data=joined, aes(x=age_z, y=age_z_std)) +
  geom_point() +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  labs(title=paste("Moran's I for dam age:",round(agecor,2)))


```

```{r}
flextable(joined |> 
  count(age_quad, sort = TRUE)|>
    st_drop_geometry())
```

### Step 5.4

**a. Spatial Clustering Patterns (Quantitative + Geographic)**

Number of dams is strongly spatially correlated. This is because dams
are concentrated in areas with wet climates, arable land, and people who
need water. Deserts and areas with low population typically do not have
large numbers of dams. Dam age is also correlated, although not as
strongly as number of dams. As the population of the US moved West, they
built dams when water resources became stretched thin between the people
living there. Most dams were also built during periods when funding was
large such as the 1930s Tennessee Valley Authority era. Dam size is not
clustered. It's unlikely a small area (such as one tile in our
tessellation) can support many large dams. Also, our metric for
investigating dam size is the mean surface area of the dams within each
grid cell. So some spatial autocorrelation is likely lost in this
averaging.

**b. Moran Scatterplot Interpretation (Visualization)**

Points generally cluster near the diagonal, suggesting positive spatial
autocorrelation. There are a few notable outliers. The number of dams is
a much stronger correlation than the age of dams. The surface area of
dams is not spatially autocorrelated.

**c. Spatial Structure Implications**

The clustering is weak for dam surface area, moderate for dam age, and
strong for number of dams. Management factors, the population expansion
of the US, and the history of dam building drives the correlation in age
and number of dams. Hydrologic factors are also visible as there are no
dams in the Southwestern deserts and there are many dams in the East.

# Question 6: Local Spatial Autocorrelation (LISA) — Mapping Hotspots and Coldspots

**In this question, you'll analyze LISA using your recommended
tessellation from Q3 AND counties as a required comparison unit.** This
contrast (regular tessellation vs administrative boundaries) reveals
MAUP much more clearly than comparing two regular grids.

### Step 6.1

```{r}
# my tesselations are already cropped and spatial lags are already calculated
```

### Step 6.2

Compute neighbors and spatial lag for your **recommended tessellation**:

```{r}
# my tesselations are already cropped and spatial lags are already calculated
```

### Step 6.3

**LOCAL Question: WHERE specifically are high-high, low-low, high-low,
and low-high clusters?** (Answer: Geographic map showing which regions
belong to which quadrant)

```{r}
ggplot(data=joined, aes(x=n_dams_z, y=n_dams_z_std)) +
  geom_point() +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0)

quad_colors <- c(
  "LL" = "blue",
  "HL" = "yellow",
  "LH" = "cyan",
  "HH" = "red"
)

ggplot() +
    geom_sf(data = joined, aes(fill = n_dams_quad)) +
    scale_color_manual(values = quad_colors) +
    theme_void() +
    labs(title = 'Number of Dams LISA plot', fill = 'Spatial correlation quadrant')
  
```

### Step 6.4

**Comparison with required county baseline**:

**a.** Repeat Steps 6.2–6.3 for counties.

**b. REQUIRED DELIVERABLE:** Create side-by-side LISA maps of **your
chosen tessellation (from Q3)** and **counties** in the same figure. Use
identical quadrant labels and color mapping in both panels so visual
differences reflect unit choice, not styling.

```{r}
library(patchwork)
left <- ggplot() +
    geom_sf(data = joined, aes(fill = n_dams_quad)) +
    scale_color_manual(values = quad_colors) +
    theme_void() +
    labs(caption = 'Raster Tesselation', fill = 'Quadrant')

right <- ggplot() +
    geom_sf(data = jc, aes(fill = n_dams_quad)) +
    scale_color_manual(values = quad_colors) +
    theme_void() +
    labs(caption = 'Counties', fill = 'Quadrant')

combo <- left + right

combo + 
  plot_annotation(
    title = "Comparing tessalations",
    theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
  )
```

**c.** **Quantitative comparison**: For both units, count the number of
polygons in each quadrant. Example output: - Hexagon: HH=15 tiles, LL=23
tiles, HL=5 tiles, LH=2 tiles - Counties: HH=205 counties, LL=1300
counties, HL=240 counties, LH=360 counties - **Interpretation**: Are
quadrant distributions similar (robust LISA pattern) or different (MAUP
sensitive at local scale)?

```{r}
flextable(joined |> 
  count(age_quad, sort = TRUE)|>
    st_drop_geometry()) |>
  set_caption('Square Grid')
```

```{r}
flextable(jc |> 
  count(age_quad, sort = TRUE)|>
    st_drop_geometry()) |>
    set_caption('Counties')
```

**d.** Interpret: Which unit reveals clustering more clearly for your
purpose? Do patterns align between your chosen tessellation and
counties, or does MAUP (modifiable areal unit problem) create divergent
local structure?

The MAUP creates different answers for each tessellation. Areas with
large counties such as Wyoming and Southern California have large
hotspots in the counties tessellation where dam density is actually
relatively low.

### Step 6.5

**Geographic Interpretation and Management Implications**

**a.** **Geographic hotspots and coldspots**: Identify major clustering
regions:

In the gridded tessellation, the hotspots are:

The great plains of central Texas, Kansas, Nebraska, and Iowa; Lots of
irrigated agriculture, medium population density The lower Mississippi
River Valley in Mississippi: Highest rainfall in the US Western and
Central Montana and South Dakota: The Missouri River and it's
tributaries are dammed extensively The Colorado Front Range: Water is
scarce, people are not The California Central Valley: Tons of irrigated
agriculture requires tons of dams Appalachia: Legacy of the TVA and
flood prevention gone overboard New England and Upstate New York: Lots
of people have been there for a long time

The county tessellation adds hotspots where there a big counties in:
Wyoming Central Washington Southern Oregon Southern California

The coldspots are:

The desert Southwest; no water, no dams Florida; Limestone geology
creates great aquifers for groundwater extraction so there is less need
for dams The upper midwest; Thousands of glacier-carved lakes create
natural reservoirs Maine; Sparsely populated and plenty of rain for
agriculture

**b.** **Outliers and anomalies**: Are there notable HL or LH tiles?
What might explain them?

HL and LH tiles typically surround coldspots and hotspots. This is just
an artifact of the LISA system and isn't very meaningful

**c.** **Unit comparison**: Does analysis unit choice (hexagon vs.
county) affect LISA conclusions? How does MAUP operate at the local
scale?

Yes, counties have different hotspots because it's not measuring dams
per unit area.

**d.** **Implications for management**: How does LISA output inform dam
operations, reporting, or policy?

I don't think the LISA output should be used to inform policy or
management in any way. In our analysis we give tiny fishing ponds formed
by a pile of dirt the same weight as Lake Mead and the Hoover Dam... Our
map loosely reflects the hydrological and anthropogenic factors that
drive dam development in the US. That is all it should be used to
inform.
